Adapting the MVDR to a GPU for active sonar systems of different sizes.

### 2013-01-22 ASA2013/asa2013.tex
la2pdf commit - 2013:01:22	18:08:51	(Tue22.Jan)

Graphics Processing Units (GPUs) is a technology tailored to perform image processing at incredible speeds. They are available off-the-shelf at reasonable prices, are very power efficient, and can be programmed to carry out a wide range of computational tasks, often displaying an order of magnitude speedups compared to similar CPU designs. Initially, GPUs may seem perfectly suited for image reconstruction algorithms, because usually each pixel can be computed independently and concurrently of one another. While this is true for simple techniques such as conventional beamformers, the more complex per-pixel computations present in adaptive techniques are harder for the GPU to handle.

%Unlike a CPU, which employ clever means to make a few threads run as fast as possible, a GPU does little to optimize the execution speed of a single thread, but in return allow thousands of them to run simultaneously.

Through a case study of on such technique, the Minimum Variance Distortionless Response (MVDR) beamformer, we identify these challenges and explain how they were overcome. The MVDR is a technique that displays a great potential for improving image quality over simple conventional methods, but is far more computationally complex. This is because a spatial covariance matrix must be estimated and inverted for each image pixel.  For systems with less than 20-30 channels, the estimation step dominates,   %Unlike other studies, w While most studies tend to focus on reducing the dimensionality of the convariance matrix to

 .

The end design is more than 2 orders of magnitude faster than a C implementation we started off with, it took considerable effort to get there.
\\
<< MOTIVATION >>
\\
Comparable can be found e.g. in ultrasound imaging, where Chen et. al have investigated porting a time domain MVDR method to a GPU \cite{Chen2009,Chen2011,Chen2011a,}. Unlike their work, however, we do not impose the restriction of the input data to be real, nor do we approach larger systems by enforcing Toeplitz structure of the covariance matrix. 
\\
This article is outlined as follows: In section \ref{methods} we will present the background on the MVDR method operated in time and frequency domain, and on the LCA beamformer. For each method we start with the mathematical formulation, then evaluate complexity issues, and finally give our thoughts on some implementational aspects. The results are collected in \ref{implementation}. Finally, we conclude in \ref{discssion}.

@@

\begin{figure}[t]\centering%
\graphicsAI[drawing,width=0.7\linewidth]{gfx/implementation.svg}%
\caption{MVDR beamforming. For one of the total number of pixels in range and azimuth, $N_y$ and $N_x$,\newline
1. an $L\times{}L$ sample covariance matrix $\eR$ is computed, \newline
2. the term $\eR^{-1}\1$ is found using a linear equation solver,\newline
3. and the beamformer output is computed from $z$ from (\ref{finalZ}), where $\w$ is found by substituting $\eR^{-1}\1$ into (\ref{weights}). }\label{mvdr_beamforming}
\end{figure}
The MVDR beamformer target this limitation by computing a set of weights for each pixel that to better fit the dynamic nature of the incoming wavefield. It computes the set of weights that minimizes the noise and signal power accumulated by the array, but with the constraint of unity gain for signal coming from the pixel of interest. When the data is delayed this is formulated as follows:
\begin{gather}
\argmin{\w}\ E\Big\{\big|\w\H[n]\x[n]\big|^2\Big\} \qquad\text{subject to}\qquad \w\H[n]\1 = 1. \label{mvdr_def}
% \vec w[n] = \frac{\Ri[n]\1}{\1\T\Ri[n]\1},\label{weights}
\end{gather}
This is a convex optimization problem with the solution:
\begin{gather}
\vec w[n] = \frac{\Ri[n]\1}{\1\T\Ri[n]\1},\label{weights}
\end{gather}
where $\1$ is a row vector of ones that represents broadside steering, and $\R=E\{\x[n]\x\H[n]\}$ is the spatial covariance matrix for the full array. Since $\R$ is unknown, we estimate it by by computing a sample covariance matrix $\eR$. In this computation we perform some degree of \emph{spatial averaging} to avoid signal cancellation, \emph{temporal averaging} to maintain true speckle statistics, and \emph{diagonal loading} to improve robustness to parameter errors ~\cite{Synnevag2009a}. Combined, these techniques will also ensure a numerically well conditioned $\eR$.

>>>>

Graphics Processing Units (GPUs) is a technology tailored to perform image processing as fast as possible. They are available off-the-shelf at reasonable prices, are very power efficient, and can be programmed to carry out a wide range of computational tasks, often displaying a speed increase of an order of magnitude compared to similar CPU designs. Initially, GPUs may seem perfectly suited for image reconstruction algorithms, because usually each pixel can be computed independently and concurrently of one another. While this is true for simple techniques such as conventional beamformers, the more complex per-pixel computations present in adaptive techniques are harder for the GPU to handle.

Through a case study of on such technique, the Minimum Variance Distortionless Response (MVDR) beamformer, we identify these challenges and explain how they were overcome. The MVDR is a technique that displays a notable potential for improving image quality over simple conventional methods, but is far more computationally complex. This is because a spatial covariance matrix must be estimated and inverted for each image pixel. For systems with less than 20-30 channels the estimation step dominates, while the inversion step   %Unlike other studies, w While most studies tend to focus on reducing the dimensionality of the convariance matrix to

The end design is more than 2 orders of magnitude faster than a C implementation we started off with, it took considerable effort to get there.
\\
<< MOTIVATION >>
\\
Comparable work can be found e.g. in ultrasound imaging, where Chen et. al have investigated porting a time domain MVDR method to a GPU \cite{Chen2009,Chen2011,Chen2011a}. Unlike their work, however, we do not impose the restriction of the input data to be real, nor do we approach larger systems by enforcing Toeplitz structure of the covariance matrix. 
\\
This article is outlined as follows: In section \ref{methods} we will present the background on the MVDR method operated in time and frequency domain, and on the LCA beamformer. For each method we start with the mathematical formulation, then evaluate complexity issues, and finally give our thoughts on some implementational aspects. The results are collected in \ref{implementation}. Finally, we conclude in \ref{discssion}.

@@

\begin{figure}[t]\centering%
\graphicsAI[drawing,width=0.7\linewidth]{gfx/implementation.svg}%
\caption{MVDR beamforming. For one of the total number of pixels in range and azimuth, $N_y$ and $N_x$,\newline
1. an $L\times{}L$ sample covariance matrix $\eR$ is computed, \newline
2. the term $\eR^{-1}\1$ is found using a linear equation solver,\newline
3. and the beamformer output is computed from $z$ from (\ref{finalZ}), where $\w$ is found by substituting $\eR^{-1}\1$ into (\ref{weights}). }\label{mvdr_beamforming}
\end{figure}
The MVDR beamformer target this limitation by adapting the weights for each pixel to best fit the dynamic nature of the incoming wavefield. It computes the set of weights that minimizes the noise and signal power accumulated by the array, but with the constraint of unity gain for signal coming from the pixel of interest. When the data is delayed this is formulated as follows:
\begin{gather}
\argmin{\w}\ E\Big\{\big|\w\H[n]\x[n]\big|^2\Big\} \qquad\text{subject to}\qquad \w\H[n]\1 = 1. \label{mvdr_def}
% \vec w[n] = \frac{\Ri[n]\1}{\1\T\Ri[n]\1},\label{weights}
\end{gather}
This is a convex optimization problem with the solution:
\begin{gather}
\vec w[n] = \frac{\Ri[n]\1}{\1\T\Ri[n]\1},\label{weights}
\end{gather}
where $\1$ is a row vector of ones that represents broadside steering, and $\R=E\{\x[n]\x\H[n]\}$ is the spatial covariance matrix for the full array. Since $\R$ is unknown, we estimate it by by computing a sample covariance matrix $\eR$. In this computation we perform some degree of \emph{spatial averaging} to avoid signal cancellation, \emph{temporal averaging} to maintain true speckle statistics, and \emph{diagonal loading} to improve robustness to parameter errors ~\cite{Synnevag2009a}. Combined, these techniques will also ensure a numerically well conditioned $\eR$.
